Query Optimization for Dynamic Imputation

### 2017-03-02 paper/introduction.tex
Minor changes to intro and motivating example

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing
values, running a sophisticated imputation algorithm over a large data set can be very
expensive---our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
queries that execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the complete imputation approach, as indicated
by one measure of error --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
percent --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

% Scope
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly deciding when to perform imputation.
% SAM -- cut the following -- could consider moving later but feels out of place here
%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches.
These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The impute operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

% Solution
One common approach is to manually replace missing values using a statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we introduce \ProjectName{}, a new system designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing
values, running a sophisticated imputation algorithm over a large data set can be very
expensive---our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
carefully plans the placement of imputation/row-drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the
Pareto-optimal frontier of plans that trade imputation cost for result quality,
and allow the analyst to specify their desired plan from this frontier.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazeroexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaoneexacs{}
percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

% Scope
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run
their queries without an explicit base-table imputation step.
% SAM -- cut the following -- could consider moving later but feels out of place here
%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches.
These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The \textit{Impute} operation fills in missing data values using any statistical imputation
  technique, such as chained-equation decision trees ~\cite{burgette2010multiple}.
  The \textit{Drop} operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the
  \textit{quality} of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-03-02 paper/main.tex
paper: more tweaks
Minor changes to intro and motivating example

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets and 
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  our experiments show that our query plans are between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the standard full-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.
\end{abstract}

>>>>

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets and 
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  our experiments show that our query plans are between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.
\end{abstract}
